Mechanising Hylomorphisms for Extraction of Executable Code

### 2025-03-24 paper/paper.tex
Containers

@@

For example, in the category \textbf{Set}, consider the polynomial functor
corresponding to
\[
F(X) = A + B \times X + C \times X^2.
\]
We can model this functor by using a set $S$ defined by $A + B + C$

which distinguishes the three cases, and a function $P$ which tells us the
number of occurrences of $X$ in each case: for $s = (1,a) \mapsto \emptyset$;
for $s = (2, b) \mapsto 1$ the singleton set and  for $s = (3, c) \mapsto 2$ the
$2$-element set.

To formalise this we need  a function $p : P \to S$ and a function
$q : P \times X \to P$.  Here, $S$ serves as a \emph{shape object} that indexes
the different cases of the functor, while $P$ represents the positions where
elements of $X$ appear. The projection $p: P \to S$ assigns each position to its
corresponding shape, and the map $q: P \to P \times X$ describes how these
positions are instantiated using elements of $X$. The functor associated with
this data is defined on objects using a dependent sum (to be precise a coend):
% \[
% F(X) = S \times_P (P \times X),
% \]
% which can also be written as a dependent sum (a coend to be precise)
\[
F(X) = \sum_{s \in S} (P(s) \to X)
\]
where  \( P(s) = p^{-1}(s) \) called the fiber of \( p \) over \( s \). Given a
function \( f: X \to Y \), the functor acts by post-composition, mapping an
element \( (s, p, x) \) in \( F(X) \) to \( (s, p, f(x)) \) in \( F(Y) \),
ensuring functoriality.

>>>>

\newcommand{\vcomp}{\circ}

@@

For example, in the category \textbf{Set}, consider the polynomial functor
corresponding to
\[
F(X) = A + B \times X + C \times X^2.
\]
We can model this functor by using a \emph{container}~\cite{AbbottAG05}. This is a set $S$
of \emph{shapes}, for example, $S = A + B + C$ for the example above, together with a function $P(s)$ denoting the exponent of $X$ for each shape.
Then a container extension is defined as the sum of all maps
$P(s) \to X$
\[
F(X) = \sum_{s \in S} (P(s) \to X)
\]
The functorial action
of $F$ is given by post-composition, mapping an element \( (s, g) \) in
\( F(X) \) to \( (s, f \vcomp g ) \) in \( F(Y) \), for $f : A \to B$.
